Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-07-11 paper-new/discussion.tex
update discussion of mechanization

\subsection{Mechanization}
% Discuss Guarded Cubical Agda and mechanization efforts
In parallel with developing the theory discussed in this paper, we have
developed a partial formalization of our results in Guarded Cubical Agda
\cite{veltri-vezzosi2020}.
%
We have formalized the key aspects in the definition of the concrete model
discussed in Section \ref{sec:concrete-model}. For instance, we have defined
types for predomains, error domains, and their morphisms and relations, and we
used the guarded features to define the guarded lift + error monad and the
dynamic type.

Building on these definitions, we formalized the notion of perturbations and
push-pull structures as well as quasi-representable relations, culminating in
the definition of value and computation types as predomains (resp. error
domains) equipped with a monoid of perturbations with an interpretation
homomorphism into the endomorphisms bisimilar to the identity. We constructed
the predomain for the dynamic type via mixed induction and guarded recursion,
defined the three relations $\inat$, $\itimes$, and $\iarr$, and proved that they are
quasi-representable.

@@

Lastly, we have formalized the adequacy result discussed in Section
\ref{sec:adequacy}. This required us to add axioms about clock quantification as
well as axioms asserting the clock-irrelevance of booleans and natural numbers,
since as of this writing these axioms are not built-in to Guarded Cubical Agda.
The mechanization of adequacy also required us to formalize some essential
lemmas involving clocks and clock-irrelevance; we may later refactor these as
part of a ``standard library'' for Guarded Cubical Agda.

>>>>

\subsection{Mechanization}
% Discuss Guarded Cubical Agda and mechanization efforts
In parallel with developing the theory discussed in this paper, we have
developed a partial formalization of our results in Guarded Cubical Agda
\cite{veltri-vezzosi2020}.
%
We have formalized the major components of the definition of the concrete model
described in Section \ref{sec:concrete-relational-model}: predomains, error
domains, their morphisms, relations, and squares, and we used the guarded
features to define the free error domain, the lock-step error ordering, and weak
bisimilarity. We also formalized the no-go theorem from Section
\ref{sec:concrete-term-model}.

Building on these definitions, we formalized the notion of semantic
perturbations and push-pull structures as well as quasi-representable relations,
culminating in the definition of value and computation types and relations as
introduced in Section \ref{sec:concrete-relational-model}.
%
% culminating in the definition of value and computation types as predomains
% (resp. error domains) equipped with a monoid of syntactic perturbations with an
% interpretation homomorphism into the semantic perturbations. 
%
We constructed the predomain for the dynamic type via mixed induction and
guarded recursion, and defined its monoid of perturbations. We also defined the
relations $\inat$, $\itimes$, and $\iarr$, and proved that they are
quasi-representable.

@@

Lastly, we have formalized the big-step term semantics discussed in Section
\ref{sec:big-step-term-semantics} and the adequacy of the relational model
discussed in Section \ref{sec:adequacy}. This required us to add axioms about
clock quantification as well as axioms asserting the \emph{clock-irrelevance} of
booleans and natural numbers, since as of this writing these axioms are not
built-in to Guarded Cubical Agda. The mechanization of adequacy also required us
to formalize some essential lemmas involving clocks and clock-irrelevance; we
are considering later refactoring these as part of a ``standard library'' for
Guarded Cubical Agda.
